---
title: Learning Genetic and Environmental Graphical Models using the R package FamilyBasedPGMs
author: Adèle Helena Ribeiro
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document: 
    keep_tex: yes
  word_document: default
  html_document: default
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


First, load the package FamilyBasedPGMs:

```{r}
library(FamilyBasedPGMs)
```

## Preparing the Dataset

Let's load the dataset containing the simulated data according to scenario 3:

```{r}
data(scen1)
```


This dataset contains 100 replicates of phenotypic data for 900 individuals (30 families, each with 30 individuals). 

The pedigrees of the families and the number of individuals in each family are in the following objects:

```{r}
pedigrees <- scen1$pedigrees
fam.nf <- scen1$fam.nf
```


The `pedigrees` object is a data.frame with the columns "famid", "id", "momid", "dadid", and "sex". The first entries for the fifth family are:

```{r echo=FALSE, results = "asis"}
stargazer::stargazer(head(subset(pedigrees, famid == 5)), summary=FALSE,
                     title = "Pedigrees", header=FALSE, label="Pedigrees")
```


To plot the pedigree chart of the fifth simulated family, you can run the following code:
 
```{r}
plotFamilyPedigree(pedigrees, famid=5)
```


In this example, let's use the first replicate of the phenotypic data:

```{r}
phen.df <- scen1$phen.df[[1]]
```


The first rows of `phen.df` are shown in the following:

```{r echo=FALSE, results = "asis"}
stargazer::stargazer(head(phen.df), summary=FALSE, 
                     title = "Phenotypes Dataset", header=FALSE,
                     label="phen.df")
```


Also, since no covariates was used in this simulation, let's define the covariate dataset as NULL:

```{r}
covs.df <- NULL
```


## Learning Total, Genetic, and Environmental Undirected PGMs

```{r echo=TRUE, cache=FALSE, results=FALSE, comment=FALSE, warning=FALSE}

scenario = 1

# Total number of individuals
N <- sum(fam.nf) 

fasterExample = TRUE

# Data was simulated for all individuals, so all individuals were "sampled".
sampled <- rep(1, N) 
if (fasterExample) {
  # If you prefer to run a faster example, you can try with a smaller part of the dataset
  # However, it may compromise the accuracy of the recovered PGMs.
  set.seed(12345)
  sampled <- sample(c(TRUE, FALSE), 900, replace=TRUE)
  phen.df <- phen.df[sampled,]
}

fileID <- paste0("scen", scenario)
dirToSave <- paste0("./objects-UDG-", fileID, "/")
dir.create(dirToSave, showWarnings=FALSE)

alpha = 0.05

udgs.out <- learnFamilyBasedUDGs(phen.df, covs.df, pedigrees, sampled, 
                                 fileID, dirToSave, alpha, correction=NULL, 
                                 max_cores=NULL, minK=10, maxFC = 0.05,
                                 orthogonal=TRUE, useGPU=FALSE, debug=TRUE)
```


Now, we can check the learned undirected *total* PGM.

Its adjacency matrix is:
```{r}
udgs.out$adjM$t
```


The estimates and p-values of the partial correlations are: 

```{r}
udgs.out$pCor$pCor_t
```


Plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
udgs.out$udg$t

plot(udgs.out$udg$t, vertex.size=30, 
     vertex.color="lightblue")

```


Let's now check the learned undirected *genetic* PGM.

Its adjacency matrix is:
```{r}
udgs.out$adjM$g
```


The estimates, p-values, and effective sizes of the partial correlations are: 

```{r}
udgs.out$pCor$pCor_g
```

Plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
udgs.out$udg$g

plot(udgs.out$udg$g, vertex.size=30, 
     vertex.color="lightblue")
```



Finally, let's check the learned undirected *environmental* PGM.

Its adjacency matrix is:
```{r}
udgs.out$adjM$e
```


The estimates, p-values, and effective sizes of the partial correlations are: 

```{r}
udgs.out$pCor$pCor_e
```


Plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
udgs.out$udg$e

plot(udgs.out$udg$e, vertex.size=30, 
     vertex.color="lightblue")
```



## Learning Total, Genetic, and Environmental Directed Acyclic PGMs

```{r echo=TRUE, cache=FALSE, results=FALSE, comment=FALSE, warning=FALSE}
fileID <- paste0("scen", scenario)
dirToSave <- paste0("./objects-PC-", fileID, "/")
dir.create(dirToSave, showWarnings=FALSE)

alpha = 0.05 # significance level

dags <- learnFamilyBasedDAGs(phen.df, covs.df, pedigrees, sampled,
   fileID, dirToSave, alpha, max_cores=NULL,
   minK=10, maxFC = 0.05, orthogonal=TRUE, maj.rule=TRUE,
   useGPU=FALSE, debug=TRUE, savePlots=FALSE)
```



Now, we can check the learned directed acyclic *total* PGM.

Its adjacency matrix is:
```{r}
adjM_t <- as(dags$t, "amat")
adjM_t 
```


Inspecting and plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
dags$t

pcalg::showEdgeList(dags$t)

igraph::plot.igraph(igraph::graph.adjacency(adjM_t), 
                    vertex.size=30, vertex.color="lightblue")
```


Let's now check the learned directed acyclic *genetic* PGM.

Its adjacency matrix is:
```{r}
adjM_g <- as(dags$g, "amat")
adjM_g 
```

Inspecting and plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
dags$g

pcalg::showEdgeList(dags$g)

igraph::plot.igraph(igraph::graph.adjacency(adjM_g), 
                    vertex.size=30, vertex.color="lightblue")
```



Finally, let's check the learned directed acyclic *environmental* PGM.

Its adjacency matrix is:
```{r}
adjM_e <- as(dags$e, "amat")
adjM_e 
```


Inspecting and plotting the learned PGM using its representation as an `igraph` object:

```{r echo=TRUE, warning=FALSE}
# igraph object
dags$e

pcalg::showEdgeList(dags$e)

igraph::plot.igraph(igraph::graph.adjacency(adjM_e),
                    vertex.size=30, vertex.color="lightblue")
```


The results of the partial correlation tests are saved at `paste0(dirToSave, fileID, "_preprPvalues.csv")`:

```{r echo=TRUE, results = "asis"}
stargazer::stargazer(round(read.table(
  paste0(dirToSave, fileID, "_preprPvalues.csv"), header=TRUE, sep=";"),
  3), summary=FALSE, title = "Partial Correlation Results", header=FALSE,
  column.sep.width = "-5pt", label="PCpcor")
```


